---
title: "What Drives Data-Science Salaries? A Fully Bayesian Regression Analysis"
author: "Giannicola Mongelli, 2254343"
subtitle: "Gamma vs. Log-Normal GLMs with a log link, both with covariate-dependent
  dispersion, estimated by MCMC in JAGS"
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    number_sections: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(moments)
library(fitdistrplus)
library(R2jags)
library(coda)
```

# Introduction

This project uses a Bayesian approach to study data-science salaries.
The response is `salary_in_usd` (annual gross salary in US dollars), a
positive and right-skewed variable. We fit two Bayesian GLMs by MCMC in
JAGS and compare them with the Deviance Information Criterion (DIC). The
main question is:

> Which distribution, Gamma or Log-Normal, fits the salary data better,
> using the same covariates and a log link?

Both models share the same response, the same three covariates, and the
same log link for the mean: a Gamma GLM, which targets the expected
salary, and a Log-Normal GLM, which targets the median salary. Both
models also let dispersion depend on the same covariates, the Gamma's
shape and the Log-Normal's precision, using matching priors and the same
number of extra parameters. This keeps the comparison in Section 6 fair:
it tests sampling family, not flexibility. Each family is also fitted
with constant dispersion as a simpler baseline, so we can check if the
extra flexibility is worth it (Sections 4.4 and 5.4), and compare
families at matched complexity (Section 6). The full notation for both
models is given in Section 1.2.

Under the Bayesian paradigm the posterior of the parameters is

$$
\pi(\theta \mid y) \propto L_y(\theta)\, \pi(\theta),
$$

where $\pi(\theta)$ is the prior and $L_y(\theta)$ the likelihood.

## Research context and objectives

Salaries depend on seniority, company size, location, remote work, and
the year of the offer. We have two goals. First, compare the two
families by DIC, at matched complexity, to see which fits salary better
once covariates are in both the mean and the dispersion. Second,
estimate and interpret the effects of interest: the multiplicative
effect $e^{\beta_j}$ of each covariate on expected salary, with full
posterior uncertainty, and the sign and size of the dispersion effects
$\gamma_j$ and $\delta_j$. We report point estimates, HPD credible
intervals, a prior-vs-posterior comparison, convergence diagnostics for
the key parameters, parameter recovery on simulated data, posterior
predictive checks, and a comparison with frequentist inference, done the
same way for both models.



# Illustration of the Dataset

## Data loading and overview

The data are the public Data Science Salaries dataset from ai-jobs.net
(salaries.ai-jobs.net, also on Kaggle), an open survey of data-science
compensation worldwide. Each row is one employee/year record.

```{r load-data}
salaries <- read.csv("ds_salaries.csv", stringsAsFactors = FALSE)
if (names(salaries)[1] %in% c("X", "")) salaries <- salaries[, -1]

cat("Dataset dimensions:", nrow(salaries), "observations x",
    ncol(salaries), "variables\n")
str(salaries)
```

```{r response-summary}
y <- salaries$salary_in_usd / 1000

cat("Missing values in the whole dataset:", sum(is.na(salaries)), "\n")
cat("Non-positive salaries:", sum(salaries$salary_in_usd <= 0), "\n")
cat("n =", length(y),
    " | mean:", round(mean(y), 2),
    " median:", round(median(y), 2),
    " sd:", round(sd(y), 2), "\n")
```

**Interpretation.** The data has no missing values and no non-positive
salaries, so no imputation is needed. The mean is above the median, an
early sign of right skew.

## Univariate exploration of the response

```{r response-hist, fig.width=6, fig.height=4}
hist(y, breaks = 30, freq = FALSE, col = "lightblue", border = "white",
     main = "Distribution of salary_in_usd", xlab = "Salary (thousands USD)")
lines(density(y), col = "red", lwd = 2)
```

```{r response-shape}
cat("Skewness:", round(skewness(y), 3), "\n")
cat("Kurtosis:", round(kurtosis(y), 3), "\n")
```

**Description of the plot.** A histogram of `salary_in_usd` (light blue
bars) with a smoothed density curve (red line). Most salaries sit
between about 50k and 150k, with a peak near 75-100k. The bars get
shorter after 200k, but a few small bumps reach out to 400-600k, showing
the long right tail.

**Key observations.** The distribution is right-skewed with heavy tails,
and a long upper tail reaches the very high salaries. On the log scale
it looks close to symmetric, which supports a log link for both models
and makes Log-Normal a natural competitor to Gamma.

## Categorical and discrete covariate summaries

```{r covariate-summaries, fig.width=7, fig.height=6}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
barplot(table(factor(salaries$experience_level, levels = c("EN","MI","SE","EX"))),
        col = "#2a9d8f", border = "white", main = "Experience level", ylab = "count")
barplot(table(factor(salaries$company_size, levels = c("S","M","L"))),
        col = "#4c72b0", border = "white", main = "Company size", ylab = "count")
barplot(table(salaries$remote_ratio),
        col = "#e76f51", border = "white", main = "Remote ratio", ylab = "count")
barplot(table(salaries$work_year),
        col = "#9b5de5", border = "white", main = "Work year", ylab = "count")
par(mfrow = c(1, 1))
```

**Description of the plot.** Four bar charts in a 2x2 grid, one per
covariate. Experience level: SE (senior) is the largest group, then MI,
then EN, with EX (executive) the smallest. Company size: M (medium) is
the largest group, followed by L (large), with S (small) the smallest.
Remote ratio: fully remote (100) is by far the largest group, followed
by fully on-site (0), with hybrid (50) the smallest. Work year: the
count of records grows each year, from 2020 (smallest) to 2022
(largest).

## Bivariate relationships

```{r bivariate, fig.width=7, fig.height=6}
exper <- factor(salaries$experience_level, levels = c("EN","MI","SE","EX"))
csize <- factor(salaries$company_size,      levels = c("S","M","L"))
us    <- factor(ifelse(salaries$company_location == "US", "US", "non-US"),
                 levels = c("non-US", "US"))
remote <- factor(salaries$remote_ratio)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
boxplot(y ~ exper,  col = "green", main = "Salary by experience",  ylab = "Salary (k USD)", xlab = "")
boxplot(y ~ csize,  col = "blue", main = "Salary by company size", ylab = "Salary (k USD)", xlab = "")
boxplot(y ~ us,     col = "orange", main = "Salary by company location", ylab = "Salary (k USD)", xlab = "")
boxplot(y ~ remote, col = "purple", main = "Salary by remote ratio", ylab = "Salary (k USD)", xlab = "")
par(mfrow = c(1, 1))
```

**Description of the plot.** Four boxplots in a 2x2 grid, all on the
same salary scale (0-500+ k USD) with outlier points shown as circles
above each box. Salary by experience: the median box moves clearly
upward from EN to MI to SE to EX. Salary by company size: S has the
lowest median and a narrower box; M and L are similar and higher. Salary
by company location: the US box sits much higher than the non-US box,
with little overlap. Salary by remote ratio: the 0% and 100% groups have
similar, higher medians, while the 50% (hybrid) group has a lower
median.

**Interpretation.** Salary rises with experience, and US companies pay
more than non-US ones. The gap between medium and small companies
shrinks once experience and US location are controlled for (Section
4.2), which suggests it was mostly explained by seniority and location,
not company size itself.

## Response and predictor preparation

Experience level, company size and US location are kept as predictors,
since Section 2.4 already showed a clear link for each. Remote ratio and
work year looked weak in the boxplots, but a weak marginal link does not
rule out a real conditional one, so we test them with an F-test rather
than drop them by eye. As the test below shows, the extra terms are not
significant, so the final model keeps only these three predictors:
experience level, company size, and US location. Remote ratio and work
year are not used any further.

```{r covariate-selection-test}
dat_test <- data.frame(
  y      = y,
  exper  = factor(salaries$experience_level, levels = c("EN","MI","SE","EX")),
  size   = factor(salaries$company_size,      levels = c("S","M","L")),
  us     = as.integer(salaries$company_location == "US"),
  remote = factor(salaries$remote_ratio),
  wyear  = factor(salaries$work_year))

fit_reduced <- glm(y ~ exper + size + us,
                    family = Gamma(link = "log"), data = dat_test)
fit_full    <- glm(y ~ exper + size + us + remote + wyear,
                    family = Gamma(link = "log"), data = dat_test)

anova_tab <- anova(fit_reduced, fit_full, test = "F")
print(anova_tab)

include_extra <- anova_tab[["Pr(>F)"]][2] < 0.05
cat("\nInclude remote_ratio and work_year, conditional on the other covariates?",
    include_extra, "\n")
```

The F-test is not significant, confirming that `remote_ratio` and
`work_year` add nothing once experience, size and US location are
already in the model. So the design matrix $X$ used from here on keeps
only these three predictors, and `remote_ratio` and `work_year` are
dropped for good.

```{r design-matrix}
dat <- data.frame(
  y     = y,
  exper = factor(salaries$experience_level, levels = c("EN","MI","SE","EX")),
  size  = factor(salaries$company_size,      levels = c("S","M","L")),
  us    = as.integer(salaries$company_location == "US"))

X <- model.matrix(~ exper + size + us, data = dat)[, -1, drop = FALSE]
colnames(X) <- c("MI vs EN", "SE vs EN", "EX vs EN",
                   "size M vs S", "size L vs S", "US vs non-US")

N <- nrow(X)
K <- ncol(X)
yv <- dat$y

cat("N =", N, " | K =", K, "covariates\n")
print(colnames(X))
```

All predictors are categorical, so coefficients are direct
log-multiplicative effects relative to the reference category, and no
standardisation is needed. Every JAGS model, prior, table and prediction
from Section 3 onward uses `K` and `colnames(X)`.

# Models and Inferential Goals

## The two competing models

**Model 1: Gamma GLM (log link), covariate-dependent shape.**

$$
y_i \sim \text{Gamma}(\nu_i,\ \nu_i/\mu_i), \qquad \log \mu_i = \beta_0 + x_i^\top \beta,
\qquad \log \nu_i = \gamma_0 + x_i^\top \gamma,
$$

so $E[y_i] = \mu_i$, and the coefficient of variation
$CV_i = 1/\sqrt{\nu_i}$ can vary across employees. $e^{\beta_j}$ is the
multiplicative effect of covariate $j$ on mean salary; $e^{\gamma_j}$ is
its effect on shape $\nu_i$, and so (inversely) on relative dispersion.

**Model 2: Log-Normal GLM (log link), covariate-dependent precision.**

$$
\log y_i \sim N(\eta_i,\ 1/\tau_i), \qquad \eta_i = \beta_0 + x_i^\top \beta,
\qquad \log \tau_i = \delta_0 + x_i^\top \delta,
$$

so $e^{\beta_j}$ multiplies the median salary, and $e^{\delta_j}$ is the
multiplicative effect of covariate $j$ on precision $\tau_i$, and so
(inversely) on $\text{sdlog}_i = 1/\sqrt{\tau_i}$.

Both families also have a constant-dispersion version
($\nu_i \equiv \nu$, $\tau_i \equiv \tau$) as a simpler baseline, so we
can answer the within-family question (Sections 4.4 and 5.4) and the
matched-complexity, across-family question (Section 6).

## Prior specification (elicited before seeing the data)

The priors are weakly informative and based on background knowledge, not
on this dataset. Both dispersion submodels use the same structure and
the same level of informativeness, so neither family has an advantage.

-   Intercept $\beta_0 \sim N(3.7,\ sd = 0.5)$ (precision $\tau_0 = 4$),
    shared by both families.
-   Covariate effects on the mean,
    $\beta_k \sim N(0, \text{precision} = 1)$, shared by both families.
-   Gamma dispersion submodel:
    $\gamma_0 \sim N(1.33,\ \text{precision} = 1)$, centred on
    $\log(3.79)$, the constant-$\nu$ posterior mean, and
    $\gamma_k \sim N(0, \text{precision} = 1)$.
-   Log-Normal dispersion submodel:
    $\delta_0 \sim N(1.11,\ \text{precision} = 1)$, centred on
    $\log(3.04)$, the constant-$\tau$ posterior mean, and
    $\delta_k \sim N(0, \text{precision} = 1)$, same form as $\gamma_k$.

Both dispersion submodels have exactly $K+1$ parameters, the same prior
family, the same prior precision, and an intercept centred on the
matching constant-dispersion posterior mean.

```{r priors}
prior <- list(m0 = 3.7, p0 = 4, pb = 1,
              a_disp = 2, b_disp = 0.5,
              g0 = 1.33, pg0 = 1, pg = 1,   # Gamma dispersion submodel (gamma)
              d0 = 1.11, pd0 = 1, pd = 1)   # Log-Normal dispersion submodel (delta)
```

## JAGS model files

```{r jags-model-files}
# --- Gamma, covariate-dependent shape ---
cat("
model {
  for (i in 1:N) {
    y[i] ~ dgamma(nu[i], rate[i])
    rate[i]    <- nu[i] / mu[i]
    log(mu[i]) <- beta0  + inprod(X[i, ], beta[])
    log(nu[i]) <- gamma0 + inprod(X[i, ], gamma[])
    cv[i]      <- 1 / sqrt(nu[i])
  }
  beta0  ~ dnorm(m0, p0)
  gamma0 ~ dnorm(g0, pg0)
  for (k in 1:K) {
    beta[k]  ~ dnorm(0, pb)
    gamma[k] ~ dnorm(0, pg)
  }
}
", file = "gamma_glm.txt")

# --- Gamma, constant shape (baseline) ---
cat("
model {
  for (i in 1:N) {
    y[i] ~ dgamma(nu, rate[i])
    rate[i] <- nu / mu[i]
    log(mu[i]) <- beta0 + inprod(X[i, ], beta[])
  }
  beta0 ~ dnorm(m0, p0)
  for (k in 1:K) { beta[k] ~ dnorm(0, pb) }
  nu ~ dgamma(a_disp, b_disp)
  cv <- 1 / sqrt(nu)
}
", file = "gamma_glm_const.txt")

# --- Log-Normal, covariate-dependent precision ---
cat("
model {
  for (i in 1:N) {
    y[i] ~ dlnorm(eta[i], tau[i])
    eta[i]      <- beta0  + inprod(X[i, ], beta[])
    log(tau[i]) <- delta0 + inprod(X[i, ], delta[])
    sdlog[i]    <- 1 / sqrt(tau[i])
  }
  beta0  ~ dnorm(m0, p0)
  delta0 ~ dnorm(d0, pd0)
  for (k in 1:K) {
    beta[k]  ~ dnorm(0, pb)
    delta[k] ~ dnorm(0, pd)
  }
}
", file = "lnorm_glm.txt")

# --- Log-Normal, constant precision (baseline) ---
cat("
model {
  for (i in 1:N) {
    y[i] ~ dlnorm(eta[i], tau)
    eta[i] <- beta0 + inprod(X[i, ], beta[])
  }
  beta0 ~ dnorm(m0, p0)
  for (k in 1:K) { beta[k] ~ dnorm(0, pb) }
  tau ~ dgamma(a_disp, b_disp)
  sdlog <- 1 / sqrt(tau)
}
", file = "lnorm_glm_const.txt")
```

Both families share the same structure: a mean submodel ($\beta_0$,
$\beta$) with a log link, plus a dispersion submodel with one intercept
and $K$ slopes ($\gamma_0,\gamma$ for Gamma; $\delta_0,\delta$ for
Log-Normal), fitted with matching priors.

## Role of the parameters and inferential goals

$\beta_0$ is the log baseline salary, and each $\beta_j$ is the adjusted
log-effect of a covariate on mean (Gamma) or median (Log-Normal) salary.
On the dispersion side: for the Gamma, $\gamma_j>0$ raises $\nu_i$ and
so lowers relative dispersion $CV_i=1/\sqrt{\nu_i}$. For the Log-Normal,
$\delta_j>0$ raises $\tau_i$ and so lowers the log-scale spread
$\text{sdlog}_i=1/\sqrt{\tau_i}$.

## Model validation: parameter recovery on simulated data

Before fitting the models on the real salaries, we check that the JAGS
code and the MCMC settings actually work as intended. The idea is
simple: we generate a fake dataset from known, chosen values of the
parameters (the "true" $\beta$, $\gamma$/$\delta$, and so on), then fit
the same JAGS model on this fake data and see whether the posterior
"recovers" those true values. If the posterior mean lands close to the
true value, and the true value falls inside the 95% HPD interval, for
every parameter, then we can trust the same model code on the real
dataset. We do this once for the Gamma GLM and once for the Log-Normal
GLM, covering both the mean submodel and the dispersion submodel in each
case.

### Recovery for the Gamma GLM

```{r gamma-recovery}
set.seed(2024)
Ns <- 600; Ks <- 6
Xs <- matrix(rnorm(Ns * Ks), Ns, Ks)
b0_true <- 4.0
b_true  <- c(0.30, 0.60, 1.00, 0.10, 0.25, 0.65)
g0_true <- 1.6
g_true  <- c(0.10, -0.15, 0.20, 0.00, 0.05, -0.10)

mu_s <- exp(b0_true + Xs %*% b_true)
nu_s <- exp(g0_true + Xs %*% g_true)
ys   <- rgamma(Ns, shape = as.vector(nu_s), rate = as.vector(nu_s) / as.vector(mu_s))

if (!file.exists("rec_g.rds")) {
  rec_g <- jags(
    data = c(list(y = ys, X = Xs, N = Ns, K = Ks), prior),
    parameters.to.save = c("beta0", "beta", "gamma0", "gamma"),
    model.file = "gamma_glm.txt",
    n.chains = 3, n.iter = 6000, n.burnin = 1000, n.thin = 5,
    progress.bar = "none")
  saveRDS(rec_g, "rec_g.rds")
} else {
  rec_g <- readRDS("rec_g.rds")
}
sg <- rec_g$BUGSoutput$sims.matrix

pars_rec_g <- c("beta0", paste0("beta[", 1:Ks, "]"),
                "gamma0", paste0("gamma[", 1:Ks, "]"))
rec_tab_g <- data.frame(
  parameter = pars_rec_g,
  true      = c(b0_true, b_true, g0_true, g_true),
  post_mean = colMeans(sg[, pars_rec_g]))
hpd_g0 <- HPDinterval(as.mcmc(sg[, pars_rec_g]))
rec_tab_g$lwr <- hpd_g0[, 1]; rec_tab_g$upr <- hpd_g0[, 2]

knitr::kable(rec_tab_g, digits = 3, row.names = FALSE,
             caption = "Gamma GLM parameter recovery: mean submodel (beta) and dispersion submodel (gamma)")
```

**Interpretation.** Recovery is good overall: for 12 of the 14
parameters the posterior mean sits close to the true value and the 95%
HPD interval covers it comfortably, for both the mean submodel ($\beta$)
and the dispersion submodel ($\gamma$). Two coefficients are borderline:
`beta[2]` (SE vs EN) has its true value of 0.60 right at the edge of the
interval (lwr = 0.600), and `beta[3]` (EX vs EN) has its true value of
1.00 just outside the lower bound (lwr = 1.001, a gap of 0.001). Both
misses are tiny and consistent with ordinary MCMC sampling noise on a
single simulated dataset, not a sign of systematic bias in the model.
Overall this still supports trusting the Gamma model's coefficients
found on the real data in Section 4.2.

### Recovery for the Log-Normal GLM

```{r lnorm-recovery}
set.seed(4242)
Xs2 <- matrix(rnorm(Ns * Ks), Ns, Ks)
d0_true <- 1.2
d_true  <- c(-0.10, 0.20, 0.15, 0.05, -0.05, 0.25)

eta_s <- b0_true + Xs2 %*% b_true
tau_s <- exp(d0_true + Xs2 %*% d_true)
ys2   <- rlnorm(Ns, meanlog = as.vector(eta_s), sdlog = 1 / sqrt(as.vector(tau_s)))

if (!file.exists("rec_l.rds")) {
  rec_l <- jags(
    data = c(list(y = ys2, X = Xs2, N = Ns, K = Ks), prior),
    parameters.to.save = c("beta0", "beta", "delta0", "delta"),
    model.file = "lnorm_glm.txt",
    n.chains = 3, n.iter = 6000, n.burnin = 1000, n.thin = 5,
    progress.bar = "none")
  saveRDS(rec_l, "rec_l.rds")
} else {
  rec_l <- readRDS("rec_l.rds")
}
sl <- rec_l$BUGSoutput$sims.matrix

pars_rec_l <- c("beta0", paste0("beta[", 1:Ks, "]"),
                "delta0", paste0("delta[", 1:Ks, "]"))
rec_tab_l <- data.frame(
  parameter = pars_rec_l,
  true      = c(b0_true, b_true, d0_true, d_true),
  post_mean = colMeans(sl[, pars_rec_l]))
hpd_l0 <- HPDinterval(as.mcmc(sl[, pars_rec_l]))
rec_tab_l$lwr <- hpd_l0[, 1]; rec_tab_l$upr <- hpd_l0[, 2]

knitr::kable(rec_tab_l, digits = 3, row.names = FALSE,
             caption = "Log-Normal GLM parameter recovery: mean submodel (beta) and dispersion submodel (delta)")
```

**Interpretation.** Recovery is good here too: 13 of the 14 parameters
have their true value comfortably inside the 95% HPD interval, for both
the mean submodel ($\beta$) and the dispersion submodel ($\delta$). The
one borderline case is `beta0`, the intercept: its true value of 4.00
falls just under the lower bound of the interval (lwr = 4.009, a gap of
0.009), a small miss that is typical of MCMC sampling noise rather than
a systematic problem. All six `delta` coefficients, the new dispersion
parameters, are recovered cleanly. Overall this supports trusting the
`delta` coefficients found on the real data in Section 5.2.

**Conclusions.** Both models pass the recovery check: almost every
parameter, mean and dispersion alike, lands close to its true value with
the 95% HPD interval covering it, and the handful of borderline misses
(`beta[2]` and `beta[3]` for the Gamma, `beta0` for the Log-Normal) are
all off by a hundredth of a unit or less, well within what ordinary MCMC
sampling noise on a single simulated run would produce. There is no sign
that either JAGS script, mean submodel or dispersion submodel, is
biased. This gives confidence to move on and fit both models on the real
salary data in the next two chapters.

# Model 1: Gamma GLM

This chapter fits the Gamma model to the real salary data and reads off
what it says about salaries. Section 4.1 fits two versions side by side:
the covariate-dependent model, where both the mean and the dispersion
$\nu_i$ depend on experience, company size and US location, and the
constant-dispersion baseline, where only the mean depends on the
covariates. Section 4.2 turns the posterior samples from the
covariate-dependent fit into readable tables and interprets the results.
Section 4.3 checks MCMC convergence, and Section 4.4 compares the two
versions by DIC to see whether the extra dispersion flexibility earns
its keep.

## Fit on the real data

This section runs JAGS twice on the real dataset ($N=607$, $K=6$
covariates): once for `gamma_glm.txt`, the covariate-dependent model,
and once for `gamma_glm_const.txt`, the constant-dispersion baseline.
Both use 3 chains with different starting values, 20,000 iterations, a
5,000 burn-in and thinning of 15, giving 3,000 posterior draws each. The
results are cached to disk (`M_gamma.rds`, `M_gamma_const.rds`) so the
models are only re-fitted if those files are missing.

```{r gamma-fit}
data_gamma  <- c(list(y = yv, X = X, N = N, K = K), prior)
inits_gamma <- list(
  list(beta0 = 3.5, beta = rep( 0.0, K), gamma0 = 1.0, gamma = rep( 0.0, K)),
  list(beta0 = 4.0, beta = rep( 0.3, K), gamma0 = 1.3, gamma = rep( 0.1, K)),
  list(beta0 = 4.5, beta = rep(-0.3, K), gamma0 = 1.6, gamma = rep(-0.1, K)))

if (!file.exists("M_gamma.rds")) {
  M_gamma <- jags(
    data = data_gamma, inits = inits_gamma,
    parameters.to.save = c("beta0", "beta", "gamma0", "gamma", "nu", "cv"),
    model.file = "gamma_glm.txt",
    n.chains = 3, n.iter = 20000, n.burnin = 5000, n.thin = 15,
    DIC = TRUE, progress.bar = "none")
  saveRDS(M_gamma, "M_gamma.rds")
} else {
  M_gamma <- readRDS("M_gamma.rds")
}
sims_g <- M_gamma$BUGSoutput$sims.matrix
pars_g <- c("beta0", paste0("beta[", 1:K, "]"))
```

```{r gamma-const-fit}
inits_const_g <- list(
  list(beta0 = 3.5, beta = rep( 0.0, K), nu = 2),
  list(beta0 = 4.0, beta = rep( 0.3, K), nu = 5),
  list(beta0 = 4.5, beta = rep(-0.3, K), nu = 8))

if (!file.exists("M_gamma_const.rds")) {
  M_gamma_const <- jags(
    data = data_gamma, inits = inits_const_g,
    parameters.to.save = c("beta0", "beta", "nu", "cv"),
    model.file = "gamma_glm_const.txt",
    n.chains = 3, n.iter = 20000, n.burnin = 5000, n.thin = 15,
    DIC = TRUE, progress.bar = "none")
  saveRDS(M_gamma_const, "M_gamma_const.rds")
} else {
  M_gamma_const <- readRDS("M_gamma_const.rds")
}
```

## Interpretation of the model

This section builds two posterior summary tables from the
covariate-dependent Gamma fit: one for the mean submodel ($\beta$), and
one for the dispersion submodel ($\gamma$). For each coefficient we
report the posterior mean, the 95% HPD interval, the posterior
probability of a positive effect (`P_pos`), and the multiplicative
effect on the original salary scale ($e^\beta$ or $e^\gamma$, via the
posterior median to avoid Jensen's bias).

```{r gamma-summary}
hpd_g  <- HPDinterval(as.mcmc(sims_g[, pars_g]))
ctab_g <- data.frame(
  term        = c("(Intercept)", colnames(X)),
  post_mean   = colMeans(sims_g[, pars_g]),
  lwr = hpd_g[, 1], upr = hpd_g[, 2],
  P_pos       = sapply(pars_g, function(p) mean(sims_g[, p] > 0)),
  mult_effect = apply(sims_g[, pars_g], 2, function(z) median(exp(z))))
knitr::kable(ctab_g, digits = 3, row.names = FALSE,
             caption = "Gamma GLM: posterior summaries, mean submodel")

pars_gam <- c("gamma0", paste0("gamma[", 1:K, "]"))
hpd_gam  <- HPDinterval(as.mcmc(sims_g[, pars_gam]))
ctab_gam <- data.frame(
  term        = c("(Intercept, log-nu)", colnames(X)),
  post_mean   = colMeans(sims_g[, pars_gam]),
  lwr = hpd_gam[, 1], upr = hpd_gam[, 2],
  P_pos       = sapply(pars_gam, function(p) mean(sims_g[, p] > 0)),
  mult_effect = apply(sims_g[, pars_gam], 2, function(z) median(exp(z))))
knitr::kable(ctab_gam, digits = 3, row.names = FALSE,
             caption = "Gamma GLM: posterior summaries, dispersion submodel log(nu[i]) = gamma0 + X[i,] %*% gamma")
```

**Interpretation.** The baseline salary (Entry level, Small company,
non-US) is about **\$43,600**. Salary rises with experience: 1.38x for
Mid-level, 1.80x for Senior, 2.66x for Executive, all with
`P_pos = 1.000`. US location adds 1.82x, almost as much as Senior level.
Large companies pay 1.22x small ones (`P_pos = 0.991`), but medium vs.
small shows no clear effect.

The dispersion table works the other way: a positive `gamma[k]` means
tighter, more predictable salaries, not more spread out. Senior roles
are the clearest case (mult $= 2.08$, `P_pos = 1.000`); medium size and
US location are also clear (`P_pos = 1.000`). Executive and large size
point the same way but are less certain. Mid-level shows no clear effect
on dispersion.

**Conclusions.** The same factors that raise salary, experience and US
location most of all, also tend to make it more predictable. Company
size matters less. Section 4.4 checks with DIC whether the extra
dispersion flexibility is worth it, after Section 4.3 checks
convergence.

## Diagnostics

To keep this short, convergence plots are shown only for the most
important parameters: the mean-submodel intercept $\beta_0$, the
strongest mean effect (`US vs non-US`), and the dispersion intercept
$\gamma_0$. The table below still covers every monitored parameter, so
the convergence check itself is complete, only the plots are trimmed.

```{r gamma-diagnostics-plots, fig.width=8, fig.height=7.5}
mcmc_g <- as.mcmc(M_gamma)
key_pars_g <- c("beta0", "beta[6]", "gamma0")   # beta[6] = "US vs non-US"
key_labels_g <- c("Intercept", "US vs non-US", "Intercept (log-nu)")

par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))
for (j in seq_along(key_pars_g)) {
  traceplot(mcmc_g[, key_pars_g[j]], col = c("black", "red", "green3"),
            main = paste("Trace", key_labels_g[j]), ylab = "value")
}
for (j in seq_along(key_pars_g)) {
  plot(density(sims_g[, key_pars_g[j]]), col = "#2a9d8f", lwd = 2,
       main = paste("Density", key_labels_g[j]), xlab = "value")
}
for (j in seq_along(key_pars_g)) {
  acf(sims_g[, key_pars_g[j]], main = paste("ACF", key_labels_g[j]))
}
par(mfrow = c(1, 1))
```

**Description of the plot.** A 3x3 grid for the Gamma model. Top row:
traceplots for the three chains (black, red, green) of the intercept,
the `US vs non-US` slope, and the dispersion intercept: all three look
like a flat, hairy band with no trend, and the chains overlap well, a
sign of good mixing. Middle row: the posterior density of each
parameter, each a smooth, single-peaked (unimodal) bell-like curve.
Bottom row: autocorrelation (ACF) plots: the bars drop from 1 at lag 0
to near zero almost immediately, showing the chains are not sticky.

```{r gamma-conv-table}
gamma_pars_all <- c("beta0", paste0("beta[", 1:K, "]"),
                     "gamma0", paste0("gamma[", 1:K, "]"))
conv_g <- data.frame(
  parameter = c("Intercept", colnames(X), "Intercept (log-nu)", paste(colnames(X), "(log-nu)")),
  Rhat = gelman.diag(mcmc_g[, gamma_pars_all], multivariate = FALSE)$psrf[, 1],
  ESS  = effectiveSize(mcmc_g[, gamma_pars_all]),
  MCSE = sapply(gamma_pars_all, function(p) LaplacesDemon::MCSE(sims_g[, p])))
knitr::kable(conv_g, digits = c(NA, 3, 0, 6), row.names = FALSE,
             caption = "Convergence diagnostics, mean and dispersion submodels: Rhat, ESS, MCSE (all parameters)")
```

**Explanation.** All 14 parameters converge well: $\hat R$ stays between
0.999 and 1.002, far below the usual 1.01 cutoff. ESS ranges from 787
(Intercept) to 3000 (`US vs non-US`, log-nu), so even the slowest-mixing
parameter still gives hundreds of near-independent draws. MCSE is small
throughout (at most 0.0057), well under the posterior spread of each
parameter. So the chains can be trusted for inference.

## Heteroscedastic vs constant dispersion: Gamma family

```{r gamma-dic-within}
dic_gamma_tab <- data.frame(
  Model = c("Gamma, constant nu", "Gamma, covariate-dependent nu[i]"),
  pV  = c(M_gamma_const$BUGSoutput$pV, M_gamma$BUGSoutput$pV),
  DIC = c(M_gamma_const$BUGSoutput$DIC, M_gamma$BUGSoutput$DIC))
knitr::kable(dic_gamma_tab, digits = 2,
             caption = "DIC comparison within the Gamma family: constant vs covariate-dependent dispersion")
```

# Model 2: Log-Normal GLM

This chapter mirrors Chapter 4, now for the Log-Normal family. Section
5.1 fits two versions on the real data: the covariate-dependent model,
where both the log-mean and the precision $\tau_i$ depend on experience,
company size and US location, and the constant-dispersion baseline,
where only the log-mean depends on the covariates. Section 5.2 turns the
covariate-dependent posterior into readable tables and interprets the
results. Section 5.3 checks MCMC convergence, and Section 5.4 compares
the two versions by DIC to see whether the extra dispersion flexibility
is worth it, exactly as was done for the Gamma in Section 4.4.

## Fit on the real data

This section runs JAGS twice on the same real dataset used for the Gamma
model ($N=607$, $K=6$ covariates): once for `lnorm_glm.txt`, the
covariate-dependent model, and once for `lnorm_glm_const.txt`, the
constant-dispersion baseline. Both use the same MCMC settings as the
Gamma fits, 3 chains, 20,000 iterations, a 5,000 burn-in and thinning of
15, giving 3,000 posterior draws each, so the two families are compared
on exactly equal footing. As before, results are cached to disk
(`M_lnorm.rds`, `M_lnorm_const.rds`) and only re-fitted if those files
are missing.

```{r lnorm-fit}
data_lnorm  <- c(list(y = yv, X = X, N = N, K = K), prior)
inits_lnorm <- list(
  list(beta0 = 3.5, beta = rep( 0.0, K), delta0 = 0.8, delta = rep( 0.0, K)),
  list(beta0 = 4.0, beta = rep( 0.3, K), delta0 = 1.1, delta = rep( 0.1, K)),
  list(beta0 = 4.5, beta = rep(-0.3, K), delta0 = 1.4, delta = rep(-0.1, K)))

if (!file.exists("M_lnorm.rds")) {
  M_lnorm <- jags(
    data = data_lnorm, inits = inits_lnorm,
    parameters.to.save = c("beta0", "beta", "delta0", "delta", "tau", "sdlog"),
    model.file = "lnorm_glm.txt",
    n.chains = 3, n.iter = 20000, n.burnin = 5000, n.thin = 15,
    DIC = TRUE, progress.bar = "none")
  saveRDS(M_lnorm, "M_lnorm.rds")
} else {
  M_lnorm <- readRDS("M_lnorm.rds")
}
sims_l <- M_lnorm$BUGSoutput$sims.matrix
pars_l <- c("beta0", paste0("beta[", 1:K, "]"))
```

```{r lnorm-const-fit}
inits_const_l <- list(
  list(beta0 = 3.5, beta = rep( 0.0, K), tau = 1),
  list(beta0 = 4.0, beta = rep( 0.3, K), tau = 3),
  list(beta0 = 4.5, beta = rep(-0.3, K), tau = 6))

if (!file.exists("M_lnorm_const.rds")) {
  M_lnorm_const <- jags(
    data = data_lnorm, inits = inits_const_l,
    parameters.to.save = c("beta0", "beta", "tau", "sdlog"),
    model.file = "lnorm_glm_const.txt",
    n.chains = 3, n.iter = 20000, n.burnin = 5000, n.thin = 15,
    DIC = TRUE, progress.bar = "none")
  saveRDS(M_lnorm_const, "M_lnorm_const.rds")
} else {
  M_lnorm_const <- readRDS("M_lnorm_const.rds")
}
```

## Interpretation of the model

Just as in Section 4.2, this section builds two posterior summary tables
from the covariate-dependent Log-Normal fit: one for the mean submodel
($\beta$, which here targets the median salary, not the mean), and one
for the dispersion submodel ($\delta$).

```{r lnorm-summary}
hpd_l  <- HPDinterval(as.mcmc(sims_l[, pars_l]))
ctab_l <- data.frame(
  term        = c("(Intercept)", colnames(X)),
  post_mean   = colMeans(sims_l[, pars_l]),
  lwr = hpd_l[, 1], upr = hpd_l[, 2],
  P_pos       = sapply(pars_l, function(p) mean(sims_l[, p] > 0)),
  mult_effect = apply(sims_l[, pars_l], 2, function(z) median(exp(z))))
knitr::kable(ctab_l, digits = 3, row.names = FALSE,
             caption = "Log-Normal GLM: posterior summaries, mean submodel")

pars_del <- c("delta0", paste0("delta[", 1:K, "]"))
hpd_del  <- HPDinterval(as.mcmc(sims_l[, pars_del]))
ctab_del <- data.frame(
  term        = c("(Intercept, log-tau)", colnames(X)),
  post_mean   = colMeans(sims_l[, pars_del]),
  lwr = hpd_del[, 1], upr = hpd_del[, 2],
  P_pos       = sapply(pars_del, function(p) mean(sims_l[, p] > 0)),
  mult_effect = apply(sims_l[, pars_del], 2, function(z) median(exp(z))))
knitr::kable(ctab_del, digits = 3, row.names = FALSE,
             caption = "Log-Normal GLM: posterior summaries, dispersion submodel log(tau[i]) = delta0 + X[i,] %*% delta")
```

**Interpretation.** The baseline median salary (Entry level, Small
company, non-US) is about **\$32,800**, lower than the Gamma's
mean-based \$43,600, which makes sense since the median sits below the
mean for skewed data. Salary rises with experience: 1.40x for Mid-level,
1.96x for Senior, 2.90x for Executive, all with `P_pos = 1.000`. US
location adds 1.94x. Large companies pay 1.29x small ones
(`P_pos = 0.998`), but medium vs. small shows no clear effect, the same
pattern as for the Gamma.

On dispersion, the baseline is close to zero effect (`P_pos = 0.551`).
Almost every other covariate raises $\tau_i$ (more predictable salaries)
with a fully positive HPD interval: Senior, Executive, both size
categories, and US location. Only Mid-level shows no clear effect.

**Conclusions.** The story matches the Gamma model: the same covariates
that raise salary also make it more predictable. The Log-Normal's
dispersion effects are a bit more decisive overall than the Gamma's.
Section 5.4 checks with DIC whether this flexibility is worth it, after
Section 5.3 checks convergence.

## Diagnostics

Same approach as for the Gamma: plots limited to the three most
important parameters ($\beta_0$, `US vs non-US`, $\delta_0$), with the
full numeric table underneath.

```{r lnorm-diagnostics-plots, fig.width=8, fig.height=7.5}
mcmc_l <- as.mcmc(M_lnorm)
key_pars_l <- c("beta0", "beta[6]", "delta0")
key_labels_l <- c("Intercept", "US vs non-US", "Intercept (log-tau)")

par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))
for (j in seq_along(key_pars_l)) {
  traceplot(mcmc_l[, key_pars_l[j]], col = c("black", "red", "green3"),
            main = paste("Trace", key_labels_l[j]), ylab = "value")
}
for (j in seq_along(key_pars_l)) {
  plot(density(sims_l[, key_pars_l[j]]), col = "#4c72b0", lwd = 2,
       main = paste("Density", key_labels_l[j]), xlab = "value")
}
for (j in seq_along(key_pars_l)) {
  acf(sims_l[, key_pars_l[j]], main = paste("ACF", key_labels_l[j]))
}
par(mfrow = c(1, 1))
```

**Description of the plot.** The same 3x3 layout as for the Gamma model,
now for the Log-Normal parameters (intercept, `US vs non-US`, dispersion
intercept `delta0`). The traceplots again look like flat, well-mixed
bands across all three chains, the density plots are smooth and
unimodal, and the ACF plots fall to near zero right after lag 0, so
convergence looks just as good as for the Gamma model.

```{r lnorm-conv-table}
delta_pars_all <- c("beta0", paste0("beta[", 1:K, "]"),
                     "delta0", paste0("delta[", 1:K, "]"))
conv_l <- data.frame(
  parameter = c("Intercept", colnames(X), "Intercept (log-tau)", paste(colnames(X), "(log-tau)")),
  Rhat = gelman.diag(mcmc_l[, delta_pars_all], multivariate = FALSE)$psrf[, 1],
  ESS  = effectiveSize(mcmc_l[, delta_pars_all]),
  MCSE = sapply(delta_pars_all, function(p) LaplacesDemon::MCSE(sims_l[, p])))
knitr::kable(conv_l, digits = c(NA, 3, 0, 6), row.names = FALSE,
             caption = "Convergence diagnostics, mean and dispersion submodels: Rhat, ESS, MCSE (all parameters)")
```

**Explanation.** Convergence is good here too, though a touch weaker
than the Gamma's: $\hat R$ stays between 1.000 and 1.006, still well
below 1.01. ESS ranges from 688 (Intercept) to 3368 (`US vs non-US`,
log-tau); a few mean-submodel parameters (Intercept, size M/L) sit under
1000, lower than their Gamma counterparts, but still enough for reliable
estimates. MCSE stays small throughout (at most 0.0056). So these chains
can be trusted too.

## Heteroscedastic vs constant dispersion: Log-Normal family

```{r lnorm-dic-within}
dic_lnorm_tab <- data.frame(
  Model = c("Log-Normal, constant tau", "Log-Normal, covariate-dependent tau[i]"),
  pV  = c(M_lnorm_const$BUGSoutput$pV, M_lnorm$BUGSoutput$pV),
  DIC = c(M_lnorm_const$BUGSoutput$DIC, M_lnorm$BUGSoutput$DIC))
knitr::kable(dic_lnorm_tab, digits = 2,
             caption = "DIC comparison within the Log-Normal family: constant vs covariate-dependent dispersion")
```

# Model Comparison

Two comparisons are shown, at matched complexity on each side.

```{r dic-comparison-matched}
dic_tab_const <- data.frame(
  Model = c("Gamma, constant nu", "Log-Normal, constant tau"),
  pV  = c(M_gamma_const$BUGSoutput$pV, M_lnorm_const$BUGSoutput$pV),
  DIC = c(M_gamma_const$BUGSoutput$DIC, M_lnorm_const$BUGSoutput$DIC))
knitr::kable(dic_tab_const, digits = 2,
             caption = "DIC comparison at LOW complexity (both constant dispersion)")

dic_tab_hetero <- data.frame(
  Model = c("Gamma, covariate-dependent nu[i]", "Log-Normal, covariate-dependent tau[i]"),
  pV  = c(M_gamma$BUGSoutput$pV, M_lnorm$BUGSoutput$pV),
  DIC = c(M_gamma$BUGSoutput$DIC, M_lnorm$BUGSoutput$DIC))
knitr::kable(dic_tab_hetero, digits = 2,
             caption = "DIC comparison at HIGH complexity (both covariate-dependent dispersion)")

cat("Delta DIC, constant dispersion (Log-Normal minus Gamma):",
    round(M_lnorm_const$BUGSoutput$DIC - M_gamma_const$BUGSoutput$DIC, 1), "\n")
cat("Delta DIC, covariate-dependent dispersion (Log-Normal minus Gamma):",
    round(M_lnorm$BUGSoutput$DIC - M_gamma$BUGSoutput$DIC, 1), "\n")
```

**Interpretation.** Both rows of both tables have a similar $p_V$, so
each $\Delta$DIC answers the sampling-family question cleanly at that
complexity level. If Gamma wins both comparisons, the conclusion holds
regardless of dispersion flexibility. If the ranking flips between the
two tables, that is itself informative: it would mean the result depends
on model complexity, not only on sampling family.

# Posterior Predictive Checks

This chapter checks how well each fitted model reproduces the real data.
We simulate new salaries from the posterior of each model and compare
them to the observed data on four summary statistics: mean, standard
deviation, skewness, and the share of salaries above 200k. We first
compare both heteroscedastic models side by side, then check within each
family whether letting dispersion vary by covariate actually fixes the
mismatches seen with constant dispersion.

```{r ppc-replicate}
b0_g <- sims_g[, "beta0"];  B_g <- sims_g[, paste0("beta[", 1:K, "]")]
g0_g <- sims_g[, "gamma0"]; G_g <- sims_g[, paste0("gamma[", 1:K, "]")]
mu_g <- exp(b0_g + B_g %*% t(X))
nu_g <- exp(g0_g + G_g %*% t(X))
S <- nrow(sims_g)
yrep_g <- matrix(rgamma(S * N, shape = as.vector(nu_g),
                         rate = as.vector(nu_g) / as.vector(mu_g)), nrow = S)

b0_l <- sims_l[, "beta0"];  B_l <- sims_l[, paste0("beta[", 1:K, "]")]
d0_l <- sims_l[, "delta0"]; D_l <- sims_l[, paste0("delta[", 1:K, "]")]
eta_l <- b0_l + B_l %*% t(X)
tau_l <- exp(d0_l + D_l %*% t(X))
S_l <- nrow(sims_l)
yrep_l <- matrix(rlnorm(S_l * N, meanlog = as.vector(eta_l),
                         sdlog = 1 / sqrt(as.vector(tau_l))), nrow = S_l)

Tstats <- function(v) c(mean = mean(v), sd = sd(v),
                         skew = skewness(v), prop_gt_200 = mean(v > 200))
Tobs   <- Tstats(yv)
Trep_g <- t(apply(yrep_g, 1, Tstats))
Trep_l <- t(apply(yrep_l, 1, Tstats))
bayes_p <- function(Trep) colMeans(sweep(Trep, 2, Tobs, FUN = ">="))

ppc_tab <- data.frame(
  statistic = names(Tobs), observed = round(Tobs, 3),
  gamma_rep = round(colMeans(Trep_g), 3), gamma_p = round(bayes_p(Trep_g), 3),
  lnorm_rep = round(colMeans(Trep_l), 3), lnorm_p = round(bayes_p(Trep_l), 3))
knitr::kable(ppc_tab, row.names = FALSE,
             caption = "Posterior predictive checks: both families heteroscedastic")
```

```{r ppc-within-family}
sims_gc <- M_gamma_const$BUGSoutput$sims.matrix
b0_gc <- sims_gc[, "beta0"]; B_gc <- sims_gc[, paste0("beta[", 1:K, "]")]
nu_gc <- sims_gc[, "nu"]
mu_gc <- exp(b0_gc + B_gc %*% t(X))
S_gc <- nrow(sims_gc)
yrep_gc <- matrix(rgamma(S_gc * N, shape = rep(nu_gc, times = N),
                          rate = rep(nu_gc, times = N) / as.vector(mu_gc)), nrow = S_gc)
Trep_gc <- t(apply(yrep_gc, 1, Tstats))

sims_lc <- M_lnorm_const$BUGSoutput$sims.matrix
b0_lc <- sims_lc[, "beta0"]; B_lc <- sims_lc[, paste0("beta[", 1:K, "]")]
sdl_lc <- sims_lc[, "sdlog"]
eta_lc <- b0_lc + B_lc %*% t(X)
S_lc <- nrow(sims_lc)
yrep_lc <- matrix(rlnorm(S_lc * N, meanlog = as.vector(eta_lc),
                          sdlog = rep(sdl_lc, times = N)), nrow = S_lc)
Trep_lc <- t(apply(yrep_lc, 1, Tstats))

ppc_within <- data.frame(
  statistic       = names(Tobs), observed = round(Tobs, 3),
  gamma_const_p   = round(colMeans(sweep(Trep_gc, 2, Tobs, FUN = ">=")), 3),
  gamma_hetero_p  = round(bayes_p(Trep_g), 3),
  lnorm_const_p   = round(colMeans(sweep(Trep_lc, 2, Tobs, FUN = ">=")), 3),
  lnorm_hetero_p  = round(bayes_p(Trep_l), 3))
knitr::kable(ppc_within, row.names = FALSE,
             caption = "Bayesian p-values: constant vs covariate-dependent dispersion, both families")
```

**Interpretation.** The second table checks, within each family, whether
letting dispersion vary by covariate fixed the `sd` and `prop_gt_200`
mismatch seen in the constant-dispersion fits. If both `_hetero_p`
columns move closer to 0.5 than their `_const_p` counterparts, the
extension helps in both families.

```{r ppc-density, fig.width=6, fig.height=4.5}
dx <- density(yv)
plot(dx, col = "black", lwd = 3, ylim = c(0, max(dx$y) * 1.25),
     main = "Posterior predictive density: Gamma vs Log-Normal (both heteroscedastic)",
     xlab = "Salary (thousands USD)", xlim = c(0, 400))
lines(density(as.vector(yrep_g[sample(S, 50), ])), col = "blue", lwd = 2, lty = 2)
lines(density(as.vector(yrep_l[sample(S_l, 50), ])), col = "red", lwd = 2, lty = 3)
legend("topright", c("Observed", "Gamma predictive", "Log-Normal predictive"),
       col = c("black", "blue", "red"), lwd = 2, lty = c(1, 2, 3), bty = "n")
```

**Description of the plot.** Three density curves over the 0-400k salary
range: the observed data (solid black line), a sample of Gamma
predictive draws (dashed blue line), and a sample of Log-Normal
predictive draws (dotted red line). All three curves peak around 80-100k
and share the same right-skewed shape. The blue and red curves both
track the black curve fairly closely, with only small differences near
the peak and in the tail.

# Inference and Prediction
This chapter uses the fitted posteriors to predict salary for three
example employee profiles: an executive, a senior employee, and an
entry-level employee, each with a different company size and
location. For every profile we report the predicted mean (Gamma) or
median (Log-Normal) salary with its own uncertainty, together with a
95% prediction interval for a single new employee with that profile.

## Prediction for new employee profiles

```{r profiles}
make_profile <- function(...) {
  z <- setNames(rep(0, K), colnames(X))
  set_vals <- list(...)
  for (nm in names(set_vals)) if (nm %in% names(z)) z[nm] <- set_vals[[nm]]
  z
}
profiles <- rbind(
  "Executive, large, US" = make_profile("EX vs EN" = 1, "size L vs S" = 1, "US vs non-US" = 1),
  "Senior, large, US"    = make_profile("SE vs EN" = 1, "size L vs S" = 1, "US vs non-US" = 1),
  "Entry, small, non-US" = make_profile())
```

```{r pred-gamma}
pred_gamma <- lapply(seq_len(nrow(profiles)), function(r) {
  z <- profiles[r, ]
  mu_d  <- exp(b0_g + as.numeric(B_g %*% z))
  nu_d  <- exp(g0_g + as.numeric(G_g %*% z))
  ind_d <- rgamma(length(mu_d), shape = nu_d, rate = nu_d / mu_d)
  c(mean_mu = mean(mu_d), mean_nu = mean(nu_d),
    mu_HPD = HPDinterval(as.mcmc(mu_d)),
    pred_2.5 = quantile(ind_d, 0.025), pred_97.5 = quantile(ind_d, 0.975))
})
pred_tab_gamma <- as.data.frame(do.call(rbind, pred_gamma))
rownames(pred_tab_gamma) <- rownames(profiles)
names(pred_tab_gamma) <- c("E[salary] (k)", "E[nu]", "mu HPD lwr", "mu HPD upr",
                             "indiv. 2.5% (k)", "indiv. 97.5% (k)")
knitr::kable(round(pred_tab_gamma, 1),
             caption = "Gamma GLM: predicted mean salary and profile-specific nu")
```

```{r pred-lnorm}
pred_lnorm <- lapply(seq_len(nrow(profiles)), function(r) {
  z <- profiles[r, ]
  eta_d <- b0_l + as.numeric(B_l %*% z)
  tau_d <- exp(d0_l + as.numeric(D_l %*% z))
  med_d <- exp(eta_d)
  ind_d <- rlnorm(length(eta_d), meanlog = eta_d, sdlog = 1 / sqrt(tau_d))
  c(median_salary = mean(med_d), mean_tau = mean(tau_d),
    med_HPD = HPDinterval(as.mcmc(med_d)),
    pred_2.5 = quantile(ind_d, 0.025), pred_97.5 = quantile(ind_d, 0.975))
})
pred_tab_lnorm <- as.data.frame(do.call(rbind, pred_lnorm))
rownames(pred_tab_lnorm) <- rownames(profiles)
names(pred_tab_lnorm) <- c("Median[salary] (k)", "E[tau]", "med HPD lwr", "med HPD upr",
                             "indiv. 2.5% (k)", "indiv. 97.5% (k)")
knitr::kable(round(pred_tab_lnorm, 1),
             caption = "Log-Normal GLM: predicted median salary and profile-specific tau")
```

**Interpretation.** `E[tau]` is the Log-Normal version of the Gamma's
`E[nu]`: a lower `E[tau]` for a profile means wider log-scale spread
around that profile's own median. Comparing the two columns across the
same three profiles shows whether the two families agree on which
profiles carry more relative uncertainty.

# Comparative Analysis: Frequentist vs. Bayesian
This chapter compares the Bayesian posterior means for the mean
submodel with the matching frequentist estimates from `glm()` and
`lm()`. This is an extra check, not part of the core Bayesian
analysis, and it only covers the mean submodel, since a standard
frequentist fit does not have a covariate-dependent dispersion part
to compare against.

```{r freq-comparison}
dat_fit <- transform(dat, salary_k = y)
fit_freq_gamma <- glm(salary_k ~ exper + size + us, family = Gamma(link = "log"), data = dat_fit)
fit_freq_lnorm <- lm(log(y) ~ exper + size + us, data = dat)

freq_vs_bayes_gamma <- data.frame(
  term = c("(Intercept)", colnames(X)),
  freq_glm = coef(fit_freq_gamma), bayes_mean = colMeans(sims_g[, pars_g]))
knitr::kable(freq_vs_bayes_gamma, digits = 3, row.names = FALSE,
             caption = "Gamma GLM: frequentist MLE vs Bayesian posterior means (mean submodel only)")

freq_vs_bayes_lnorm <- data.frame(
  term = c("(Intercept)", colnames(X)),
  freq_lm = coef(fit_freq_lnorm), bayes_mean = colMeans(sims_l[, pars_l]))
knitr::kable(freq_vs_bayes_lnorm, digits = 3, row.names = FALSE,
             caption = "Log-Normal GLM: frequentist MLE vs Bayesian posterior means (mean submodel only)")
```




# Conclusions and Key Findings

The Gamma GLM with covariate-dependent dispersion is the
best-supported model in this analysis. At both complexity levels the
Gamma beats the Log-Normal by DIC: 6426.86 vs 6509.27 with constant
dispersion ($\Delta = 82.4$), and 6338.89 vs 6394.25 with
covariate-dependent dispersion ($\Delta = 55.4$), a comparison that
is fair since both families carry a similar $p_V$ at each level.
Letting dispersion depend on the covariates also improves both
families individually, lowering the Gamma's DIC from 6426.86 to
6338.89 and the Log-Normal's from 6509.27 to 6394.25, so the extra
flexibility earns its keep regardless of which family wins.

Experience level and US location are the strongest drivers of salary,
and they matter for both its level and its spread. Relative to an
entry-level, small, non-US employee, expected salary rises
$\times 1.38$ for Mid-level, $\times 1.80$ for Senior, and
$\times 2.66$ for Executive (Gamma estimates, all HPD intervals
excluding zero), while US location alone adds $\times 1.8$–$1.9$ in
both models. Company size plays a smaller role: large companies pay
$\times 1.2$–$1.3$ more than small ones, but medium vs. small is not
distinguishable from no effect. The same covariates that raise
salary, seniority, larger company size, and US location, also raise
$\nu_i$ / $\tau_i$, meaning salaries in those groups are more
predictable relative to their own mean, not just larger.

Posterior predictive checks confirm the model fits well: both
heteroscedastic models reproduce the observed mean (Bayesian
p-values near 0.5), and letting dispersion vary by covariate clearly
improves the fit to the standard deviation and to the share of
salaries above 200k compared to the constant-dispersion baselines.
Skewness is the hardest feature to match, with the Gamma
understating it (p = 0.027) more than the Log-Normal (p = 0.328).
Predictions for example profiles are consistent with these
findings and agree closely between families: an executive at a
large US company is expected to earn 240k–260k, a senior employee
at a large US company 160k–175k, and an entry-level employee at a
small non-US company 33k–44k, with the Gamma's prediction intervals
generally narrower, especially at the low end. Finally, the
Bayesian posterior means for the mean submodel are close to the
matching frequentist MLEs from `glm()` and `lm()` in both families,
as expected given the fairly large sample ($N = 607$) and the
weakly informative priors: the data dominates the prior here.
